Local Type Inference for Polarised System F with Existentials

### 2024-05-07 oopsla24/tex/05-proof.tex
Edit: sections 1-5

\label{sec:proofs}
The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.
The soundness means the algorithm will always
produce a typing \emph{allowed} by the declarative system;
Dually, the completeness says that once a term has some type declaratively, the inference algorithm succeeds.

The formulation of soundness and completeness of \emph{application inference} is more complex.
Both of them assume that the input head type $[[uN]]$ is free from \emph{negative}
algorithmic variables (it is achieved by polarization invariants preserved by the inference rules).
The soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$ is viable. 
Specifically, that the constraint set $[[SC]]$ provides a sufficient set of restrictions that 
a substitution $[[uσ]]$ must satisfy to ensure the \emph{declarative} inference of the output type $[[uM]]$, that is 
$[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$.

@@

\footnotetext[1]{The presented properties hold, 
            but the actual inductive proof requires strengthening 
            of the statement and the corresponding theorem is more involved. 
            See the appendix for details.}

\begin{theorem*}[Soundness of Typing]
    \label{thm:soundness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \hfill
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$,
            assuming $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
            \emph{negative} algorithmic variables.
    \end{itemize}
\end{theorem*}

@@

\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, the correctness of normalization means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization Correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

@@

The anti-unifier of the two types is the most specific pattern that matches against both of them. 
In our setting, the anti-unifiers are restricted further: first, the pattern might only contain placeholders at 
\emph{negative} positions (because eventually, the placeholders become variables abstracted by the existential quantifier);
second, the anti-unification is parametrized with a context $[[Γ]]$, in which the pattern instantiations must be well-formed.

This way, the correctness properties of the anti-unification algorithm are refined accordingly.
The soundness of anti-unification not only ensures that the resulting pattern matches with 
the input types, but also that the pattern instantiations are well-formed in the corresponding context, and 
that all the `placeholder' variables are negative. In turn, completeness states that if there exists a solution 
satisfying the soundness criteria, then the algorithm succeeds.

@@

\begin{lemma*}[Soundness of (Positive) Anti-Unification]
    \label{lemma:au-soundness}
     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
    if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
    then 
    \begin{enumerate*}
        \item $[[Γ ; Ξ ⊢ uQ]]$,
        \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
        \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}
\end{lemma*}

\begin{lemma*}[Completeness of (Positive) Anti-Unification]
    \label{lemma:au-completeness}
    Assuming that $[[iP1]]$ and $[[iP2]]$ are normalized terms well-formed in $[[Γ]]$
    and there exist  $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
    \begin{enumerate*}
        \item $[[Γ ; Ξ' ⊢ uQ']]$,
        \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
        are anti-unification substitutions, and
        \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}

@@

The anti-unification is used to find the least upper bound (LUB).
To guarantee that the result of the LUB algorithm is indeed the least, 
we also need to show the corresponding property of anti-unification.
We prove that the anti-unifier that the algorithm provides 
is the most specific (or the most `detailed'). Specifically, 
it means that any other sound anti-unification solution 
can be `refined' to the result of the algorithm.  
The `refinement' is represented as an instantiation of the anti-unifier---a 
substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the placeholders $[[Ξ1]]$ with
types that themselves can contain placeholders from $[[Ξ2]]$.

@@

To prove the correctness properties of the anti-unification algorithm,
one extra observation is essential. The algorithm relies on the fact that
in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
there are no two different placeholders $[[β̂⁻]]$
mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
This is important to guarantee that, for example, 
the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
(less specific) $[[↓â⁻ → b̂⁻]]$.
To preserve this property, we arrange the algorithm in such a way 
that the name of the placeholder is determined by the types it is mapped to.
The following lemma specifies this observation.

@@

    The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
    The \emph{soundness} means that the returned type is indeed a supertype of the given ones;
    the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
    and the \emph{initiality} means that the returned type is the least among common supertypes.

    \begin{lemma*}[Least Upper Bound Soundness]
        For types $[[Γ ⊢ iP1]]$, and $[[Γ ⊢ iP2]]$,
        if $[[Γ ⊨ iP1 ∨ iP2 = iQ]]$ then
        \begin{enumerate}
            \item[(i)]  $[[Γ ⊢ iQ]]$
            \item[(ii)] $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$
        \end{enumerate}
    \end{lemma*}

    \begin{lemma*}[Least Upper Bound Completeness and Initiality]
        For types $[[Γ ⊢ iP1]]$, $[[Γ ⊢ iP2]]$, and $[[Γ ⊢ iQ]]$
        such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
        there exists $[[iQ']]$ s.t. $[[Γ ⊨ iP1 ∨ iP2 = iQ']]$ 
        and $[[Γ ⊢ iQ ≥ iQ']]$.
    \end{lemma*}

    The key observation that allows us to prove these properties is
    the characterization of positive supertypes. 
    The following lemma justifies the cases of the Least Upper Bound algorithm
    (\cref{sec:lub}). In particular, it establishes
    the correspondence between the upper bounds of shifted types $[[↓iM]]$ and
    \emph{patterns} fitting $[[iM]]$ (represented by existential types), which 
    explains the usage of anti-unification as a way to find a common pattern.

    \begin{lemma*}[Characterization of Normalized Supertypes]
        \label{lemma:char-supertypes}
        For a normalized positive type $[[iP]]$ well-formed in $[[Γ]]$,
        the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
        \begin{itemize}
            \item if $[[iP]]$ is a variable $[[pb]]$, its only normalized supertype is $[[pb]]$ itself;
            \item if $[[iP]]$ is an existential type $[[ ∃nbs.iP' ]]$ then 
                its $[[Γ]]$-formed supertypes are the $([[G, nbs]])$-formed supertypes of $[[iP']]$ not using $[[nbs]]$;
            \item if $[[iP]]$ is a downshift type $[[↓iM]]$, 
                its supertypes have form $[[∃nas.↓iM']]$ such that there exists
                a $[[Γ]]$-formed instantiation of $[[nas]]$ in $[[↓iM']]$
                that makes $[[↓iM']]$ equal to $[[↓iM]]$, \ie $[[ [iNs/nas] ↓iM' = ↓iM ]]$.
        \end{itemize}
    \end{lemma*}

@@

    \begin{lemma*}[Upgrade Soundness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
        then
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ]]$
            \item $[[Γ ⊢ iQ ≥ iP]]$
        \end{enumerate*}
    \end{lemma*}

    \begin{lemma*}[Upgrade Completeness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        for any $[[iQ']]$ such that $[[iQ']]$ is a $[[Δ]]$-formed upper bound of $[[iP]]$, \ie
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ']]$ and
            \item $[[Γ ⊢ iQ' ≥ iP]]$,
        \end{enumerate*}
        there exists$[[iQ]]$ such that
        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and $[[Δ ⊢ iQ' ≥ iQ]]$.
    \end{lemma*}

@@

    As for other properties, the correctness of subtyping means that
    the algorithm produces a valid result (soundness) 
    whenever it exists (completeness).
    In \cref{fig:alg-subtyping}, one can see that the positive and 
    negative subtyping relations are not \emph{mututally} recursive: 
    negative subtyping algorithm uses the positive subtyping, 
    but not vice versa. 
    Because of that, the inductive proofs of the \emph{positive} subtyping correctness are done independently.

    The soundness of positive subtyping states that
    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
    In other words, any substitution satisfying $[[SC]]$, 
    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
    Notice that the soundness requires that only the left-hand side input 
    type ($[[uP]]$) can have algorithmic variables. This is one of the invariants of the algorithm
    that significantly simplifies the unification and constraint resolution.

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$ and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

    The completeness of subtyping says that if the substitution 
    ensuring the declarative subtyping exists, then the algorithm terminates.

\begin{lemma*}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv(uP)]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    then the subtyping algorithm succeeds: $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
\end{lemma*}

After the correctness properties of the positive subtyping are established,
they are used to prove the correctness of the negative subtyping.
The soundness is formulated symmetrical to the positive case,
however, the completeness requires an additional invariant to be preserved.
The algorithmic input type $[[uN]]$ must be free from \emph{negative} algorithmic variables.
In particular, it ensures that the constraint restricting a \emph{negative} algorithmic
variable will never be generated, and thus, we do not need the 
Greatest Common Subtype procedure to resolve the constraints.

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iM]]$ and $[[Γ ; dom(Θ) ⊢ uN]]$
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
\end{lemma*}

@@

\paragraph{Minimal Instantiation}
Algorithmic typing relies on the \emph{minimal instantiation} procedure.
For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
This way, the correctness of minimal instantiation is formulated as the following two lemmas.

\begin{lemma*}[Soundness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
    Then $[[Γ ⊢ uP SC minby uσ ]]$ implies that 
    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
    (\ie $[[Θ ⊢ uσ : SC]]$)
    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$ holds
\end{lemma*}

\begin{lemma*}[Completeness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
    satisfying $[[SC]]$, we have
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
    succeeds, specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
\end{lemma*}

@@

\paragraph{Singularity}
The soundness of singularity states that $[[SC singular with uσ]]$
implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
on the domain.  The completeness states that if all the 
$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.

@@

\begin{lemma*}[Completeness of Singularity]
    \label{lemma:singularity-completeness}
    For a given $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : dom(SC)]]$ such that
    for any $[[Θ ⊢ uσ : dom(SC)]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : dom(SC)]]$.
    Then\footnotemark[1] $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
\end{lemma*}

@@

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires certain properties of the premise subtrees.
However, in the case of the typing algorithm, the soundness and completeness proofs
cannot be separated: the inductive proof of one requires another, and vice versa.

The soundness proof can be viewed as a mapping from an algorithmic tree to a
declarative one. We show that each algorithmic inference rule can
be transformed into the corresponding declarative rule, as long as the premises
are transformed accordingly, and apply the induction principle.

>>>>

\label{sec:proofs}
The central results ensuring the correctness of the inference algorithm are its
soundness and completeness with respect to the declarative specification. The
soundness means the algorithm will always produce a typing \emph{allowed} by the
declarative system; dually, the completeness says that once a term has some type
declaratively, the inference algorithm succeeds.

The formal statements of soundness and completeness are given in the theorems
below. Notice that the theorems also include the soundness and completeness of
\emph{application inference} (labeled as $\bullet$), which is more complex. As
such, let us discuss it in more detail.

Both soundness and completeness of application inference assume that the input
head type $[[uN]]$ is free from \emph{negative} algorithmic variables---it is
achieved by polarization invariants preserved by the inference rules. The
soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$---is
viable. Specifically, that the constraint set $[[SC]]$ provides a sufficient set
of restrictions that any substitution $[[uσ]]$ must satisfy to ensure the
\emph{declarative} inference of the output type $[[uM]]$, that is 
$[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$.

@@

\footnotetext[1]{The presented properties hold, 
            but the actual inductive proof requires strengthening 
            of the statement and the corresponding theorem is more involved. 
            See the appendix (\cref{sec:appendix}) for details.}

\begin{theorem*}[Soundness of Typing]
    \label{thm:soundness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \hfill
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$.
            All of it under assumptions that $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
            \emph{negative} algorithmic variables.
    \end{itemize}
\end{theorem*}

@@

\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, \emph{the correctness of normalization} means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

@@

The anti-unifier of the two types is the most specific pattern that matches
against both of them. In our setting, the anti-unifiers are restricted further:
first, the pattern might only contain `holes' at \emph{negative} positions
(because eventually, the `holes' become variables abstracted by the
existential quantifier); second, the anti-unification is parametrized with a
context $[[Γ]]$, in which the pattern instantiations must be well-formed.

This way, the correctness properties of the anti-unification algorithm are
refined accordingly. The soundness of anti-unification not only ensures that the
resulting pattern matches with the input types, but also that the pattern
instantiations are well-formed in the corresponding context, and that all the
`hole' variables are negative. In turn, completeness states that if there
exists a solution satisfying the soundness criteria, then the algorithm
succeeds.

@@

\begin{lemma*}[Soundness of (positive) anti-unification]
    \label{lemma:au-soundness}
     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
    if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
    then 
    \begin{enumerate*}
        \item $[[Γ ; Ξ ⊢ uQ]]$,
        \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
        \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}
\end{lemma*}

\begin{lemma*}[Completeness of (positive) anti-unification]
    \label{lemma:au-completeness}
    Assuming that $[[iP1]]$ and $[[iP2]]$ are normalized terms well-formed in $[[Γ]]$
    and there exist  $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
    \begin{enumerate*}
        \item $[[Γ ; Ξ' ⊢ uQ']]$,
        \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
        are anti-unification substitutions, and
        \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
    \end{enumerate*}

@@

Notice that for the anti-unification substitution $[[aus]]$ we use a separate
signature specifying the domain and the range. $[[Γ ; Ξ2 ⊢ aus : Ξ1]]$ means
that $[[aus]]$ maps the `holes' (\ie algorithmic variables) from $[[Ξ1]]$
to \emph{algorithmic} types well-formed in $[[Γ]]$ and $[[Ξ2]]$. To put it
formally, $[[Γ ; Ξ2 ⊢ [aus]α̂⁻]]$ for any $[[α̂⁻ ∊ Ξ1]]$.
Although in the formulation of soundness and completeness, the resulting types are declarative
(\ie $[[Ξ2]]$ is always empty), we need the possibility of mapping the
`holes' to types with `holes' to formulate the \emph{initiality} of
anti-unification.

The initiality shows that the anti-unifier that the algorithm provides is the
most specific (or the most `detailed'). Specifically, it means that any other
sound anti-unification solution can be `refined' to the result of the algorithm.  
The `refinement' is represented as an instantiation of the anti-unifier---a 
substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the `holes' from $[[Ξ1]]$ with
types that themselves can contain `holes' from $[[Ξ2]]$.

@@

To prove the correctness properties of the anti-unification algorithm,
one extra observation is essential. The algorithm relies on the fact that
in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
there are no two different `holes' $[[β̂⁻]]$
mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
This is used to ensure that, for example, 
the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
(less specific) $[[↓â⁻ → b̂⁻]]$.
To preserve this property, we arrange the algorithm in such a way 
that the name of the `hole' is determined by the types it is mapped to.
The following lemma specifies this observation.

@@

    The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
    Its \emph{soundness} means that the returned type is indeed a supertype of the given ones;
    the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
    and the \emph{initiality} means that the returned type is the least among common supertypes.

    \begin{lemma*}[Least Upper Bound soundness]
        For types $[[Γ ⊢ iP1]]$, and $[[Γ ⊢ iP2]]$,
        if $[[Γ ⊨ iP1 ∨ iP2 = iQ]]$ then
        \begin{enumerate}
            \item[(i)]  $[[Γ ⊢ iQ]]$
            \item[(ii)] $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$
        \end{enumerate}
    \end{lemma*}

    \begin{lemma*}[Least Upper Bound completeness and initiality]
        For types $[[Γ ⊢ iP1]]$, $[[Γ ⊢ iP2]]$, and $[[Γ ⊢ iQ]]$
        such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
        there exists $[[iQ']]$ s.t. $[[Γ ⊨ iP1 ∨ iP2 = iQ']]$ 
        and $[[Γ ⊢ iQ ≥ iQ']]$.
    \end{lemma*}

    The key observation that allows us to prove these properties is the
    characterization of positive supertypes. The following lemma justifies the
    case analysis used in the Least Upper Bound algorithm (in \cref{sec:lub}).
    In particular, it establishes the correspondence between the upper bounds of
    shifted types $[[↓iM]]$ and \emph{patterns} fitting $[[iM]]$ (represented by
    existential types), which explains the usage of anti-unification as a way to
    find a common pattern.

    \begin{lemma*}[Characterization of normalized supertypes]
        \label{lemma:char-supertypes}
        For a normalized positive type $[[iP]]$ well-formed in $[[Γ]]$,
        the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
        \begin{itemize}
            \item if $[[iP]]$ is a variable $[[pb]]$, its only normalized supertype is $[[pb]]$ itself;
            \item if $[[iP]]$ is an existential type $[[ ∃nbs.iP' ]]$ then 
                its $[[Γ]]$-formed supertypes are the $([[G, nbs]])$-formed supertypes of $[[iP']]$ not using $[[nbs]]$;
            \item if $[[iP]]$ is a downshift type $[[↓iM]]$, 
                its supertypes have form $[[∃nas.↓iM']]$ such that there exists
                a $[[Γ]]$-formed instantiation of $[[nas]]$ in $[[↓iM']]$
                that makes $[[↓iM']]$ equal to $[[↓iM]]$, \ie $[[ [iNs/nas] ↓iM' = ↓iM ]]$.
        \end{itemize}
    \end{lemma*}

@@

    \begin{lemma*}[Upgrade soundness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
        then
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ]]$ and
            \item $[[Γ ⊢ iQ ≥ iP]]$.
        \end{enumerate*}
    \end{lemma*}

    \begin{lemma*}[Upgrade Completeness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        for any $[[iQ']]$ such that $[[iQ']]$ is a $[[Δ]]$-formed upper bound of $[[iP]]$, \ie
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ']]$ and
            \item $[[Γ ⊢ iQ' ≥ iP]]$,
        \end{enumerate*}
        there exists $[[iQ]]$ such that
        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and $[[Δ ⊢ iQ' ≥ iQ]]$.
    \end{lemma*}

@@

    As for other properties, the correctness of subtyping means that the
    algorithm produces a valid result (soundness) whenever it exists
    (completeness). In the rules defining the subtyping algorithm
    (\cref{fig:alg-subtyping}), one can see that the positive and negative
    subtyping relations are not \emph{mututally} recursive: negative subtyping
    algorithm uses the positive subtyping, but not vice versa. Because of that,
    the inductive proofs of the \emph{positive} subtyping correctness are done
    independently.

    The soundness of positive subtyping states that the output constraint
    $[[SC]]$ provides a sufficient set of restrictions. In other words, any
    substitution satisfying $[[SC]]$ ensures the desired declarative subtyping:
    $[[Γ ⊢ [uσ]uP ≥ iQ]]$. Notice that the soundness formulation assumes that
    only the left-hand side input type ($[[uP]]$) can contain algorithmic
    variables. This is one of the invariants of the algorithm that significantly
    simplifies the unification and constraint resolution.

\begin{lemma*}[Soundness of Positive Subtyping]
    If $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$ and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

    The completeness of subtyping says that if the substitution 
    ensuring the declarative subtyping exists, then the subtyping algorithm 
    succeeds (terminates).

\begin{lemma*}[Completeness of Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv(uP)]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    then the subtyping algorithm succeeds: $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
\end{lemma*}

After the correctness properties of positive subtyping are established,
they are used to prove the correctness of the negative subtyping.
The soundness is formulated symmetrical to the positive case,
however, the completeness requires an additional invariant to be preserved.
The algorithmic input type $[[uN]]$ must be free from \emph{negative} algorithmic variables.
In particular, it ensures that the constraint restricting a \emph{negative} algorithmic
variable will never be generated, and thus, we do not need the 
Greatest Common Subtype procedure to resolve the constraints.

\begin{lemma*}[Completeness of Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iM]]$ and $[[Γ ; dom(Θ) ⊢ uN]]$
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
\end{lemma*}

@@

Algorithmic typing relies on the \emph{minimal instantiation} procedure.
For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
This way, the correctness of minimal instantiation is formulated as the following two lemmas.

\begin{lemma*}[Soundness of Minimal Instantiation]
    \label{lemma:minimal-instantiation-soundness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
    Then `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
    (\ie $[[Θ ⊢ uσ : SC]]$)
    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
    we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
    satisfying $[[SC]]$, we have
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
\end{lemma*}

@@

The soundness of \emph{singularity} states that $[[SC singular with uσ]]$
implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
on the domain.  The completeness states that if all the 
$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.

@@

\begin{lemma*}[Completeness of Singularity]
    \label{lemma:singularity-completeness}
    For a given set of constraints $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : dom(SC)]]$ such that
    for any $[[Θ ⊢ uσ : dom(SC)]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : dom(SC)]]$.
    Then $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
\end{lemma*}

@@

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires the premise subtrees to satisfy certain properties
(which can be given by the soundness). However, in the case of the typing algorithm,
the soundness and completeness proofs cannot be separated: the inductive proof
of one requires another, and vice versa.

The proof of soundness can be viewed as a mapping from an algorithmic tree to a
declarative one. We show that each algorithmic inference rule can be transformed
into the corresponding declarative rule, as long as the premises are transformed
accordingly, and apply the induction principle.
